Developmental trajectories of bilingual word recognition

### 2023-11-22 manuscript/_appendix/_stimuli.qmd
Move stimuli details to appendix

---
format:
  docx: default
---

```{r setup}
#| label: setup
#| echo: false
#| message: false
#| warning: false
# load objects
targets::tar_config_set(store = here::here("_targets"),
						script = here::here("_targets.R"))

targets::tar_load_globals()

targets::tar_load(
	c(participants,
	  stimuli,
	  vocabulary,
	  attrition_trials,
	  attrition_participants,
	  bvq_data,
	  gaze,
	  data_bcn,
	  data_oxf
	))

# targets::tar_load(
# 	c(model_fits_bcn,
# 	  model_loos_bcn,
# 	  model_fits_oxf,
# 	  model_loos_oxf
# 	))

# prepare data
attrition <- lst(participants,
				 attrition_participants,
				 attrition_trials) |>
	reduce(inner_join, by = join_by(session_id)) |>
	unnest_wider(matches("all"), names_sep = "_") |> 
	rename_with(\(x) gsub("is_valid_gaze_all_|is_valid_vocab_all", "", x))

library(knitr)
library(kableExtra)
library(ggplot2)
library(gt)
library(gtExtras)
library(patchwork)
library(english)
library(gtExtras)

# set ggplot theme and colour palette
my_theme <- theme_minimal() +
	theme(panel.grid = element_blank(),
		  axis.line = element_line(colour = "black"),
		  text = element_text(size = 12, colour = "black"),
		  axis.text = element_text(colour = "black"))

theme_set(my_theme)

clrs <- c("#003f5c", "#58508d", "#bc5090", "#ff6361", "#ffa600")

options(ggplot2.ordinal.fill = clrs[c(1, 4, 5)],
		ggplot2.ordinal.colour = clrs[c(1, 4, 5)],
		ggplot2.discrete.fill = clrs[c(1, 4, 5)],
		ggplot2.discrete.colour = clrs[c(1, 4, 5)],
		ggplot2.continuous.fill = ggplot2::scale_color_gradient,
		ggplot2.continuous.colour = ggplot2::scale_color_gradient)

set.seed(1234)
```

# Introduction

Building a mental lexicon is a major achievement in the development of an infant: by storing representations of how familiar words sound and what they mean, an infant is able to make sense of their linguistic input. The foundations of an initial lexicon are in place before their first year of life [@tincoff1999some; @halle1994emergence; @vihman2004cross; @parise2012electrophysiological; @bergelson20126; @bergelson2015early]. This initial lexicon consists of only a few items; mainly words for people, interjections, body parts, and food [@tincoff2012six; @tardif2008baby], but it undergoes rapid growth during the second year of life [@ganger2004reexamining; @goldfield1990early; @bloom2002children; @mcmurray2007defusing]. According to parental reports, the average 15-month-old infant already understands more than 100 words, and by two years, they understand more than 400 words [@frank2021variability]. This accelerated lexical developmental is reflected in infants' trajectories of word recognition: infants recognise familiar words faster and more efficiently as they approach their second birthday [@fernald1998rapid; @fernald2001half; @hurtado2007spoken].

Despite being exposed to a more complex linguistic input, bilinguals show equivalent trajectories of word acquisition and word recognition to their monolingual peers' [@vihman2007onset; @legacy2018vocabulary; @hoff2012dual; @de2014bilingual; @pearson1994patterns; @byers-heinlein2023sometimes]. This is a remarkable deed for two reasons. First, bilingual infants receive a relative impoverished linguistic input in each of their languages compared to monolinguals [@costa2014does]. Second, they face a more complex referential context, since they often learn two labels for each referent, one in each language. The mechanisms that allow bilingual' trajectories of lexical developmental to keep up with monolinguals' are still unclear.

Monolinguals and bilinguals share broadly common word-recognition mechanisms. The first step towards spoken word-recognition is the acoustic-phonetic processing of the speech signal, which results in the activation of lexical representations whose associated phonological forms match the acoustic signal. The activated word-forms then compete for selection: the lexical system uses the bottom-up information provided by the auditory input, together with top-down context---provided by grammatical, supra-segmental, and semantic constraints---to select the candidate lexical representation with the best match. This sequence of events occurs in a *cascaded* fashion. As the recognition system accumulates information about the unfolding acoustic signal, matching phonological representations are activated [@marslen1987functional; @marslen1978processing; @grosjean1980spoken], and in turn activate their associated semantic representations [@neely1977semantic]. By the time lexical selection occurs, many non-selected phonological and semantic representations have been activated, potentially modulating the dynamics of word recognition.

A considerable amount of evidence for such cascaded account of lexical processing has been provided by studies recording eye movements during word recognition. For example, @allopenna1998tracking designed a word-recognition task in which participants were presented, in each trial, with four objects in the screen. Participants would then listen to a command like "Pick up the beaker; now put it below the diamond", and perform the action. The authors manipulated the phonological relationship between the target object's label (e.g., *casket*), and the label of each of the three distractors. Two of the distractors were phonologically related to the target, sharing onset (e.g., *castle*) or offset (e.g., *basket*), respectively. The other distractor was phonologically unrelated to the target (e.g., *nickel*). After the onset of the auditory target label, phonological distractors attracted participants' eye fixations more than unrelated distractors. Distractors sharing phonological onset did so at an earlier time window than those sharing phonological offset. This time course of distractor fixations suggests that the auditory presentation of the target label activated no only its (subsequently selected) phonological form in participants' lexicon, but also that of phonologically related word-forms, leading to competition for selection between the word forms. These results converge with cascaded models of auditory word recognition, in which the unfolding acoustic input activates multiple phonological word-forms, which in turn compete for selection, like TRACE [@mcclelland1986trace] and Shortlist [@norris1994shortlist].

The developing mental lexicon also seems to operate in this cascaded fashion: like adults, infants activate non-selected lexical representations. For instance, priming paradigms of word recognition have provided a strong body of evidence for the existence of phonological and semantic links between early lexical representations between 18 and 21 months of age.. @arias2009lexical found that infants' spoken word recognition (e.g., *dot*) was interfered by the previous presentation of a semantically related word (e.g., *cat*). Other studies have provided electrophysiological evidence of semantic priming at earlier ages [e.g., @rama2013development; @friedrich2005lexical]. More recent studies have provided evidence for the emergence of such semantic links, even in the absence of visual referents during the experimental task [@willits2013toddlers]. Phonological priming effects have also been reported in the initial lexicon. @mani2010infant and @mani2011phonological created an implicit naming task in which each trial started with the *silent* presentation of a prime picture. Then, a target-distractor picture pair was presented side-by-side. Finally, the target picture's label was presented auditorily. The authors manipulated the phonological overlap between the prime label and the target label, so that in half of the trials both labels were phonologically related, sharing phonological onset (*cat*-*cup*), or phonologically unrelated (*ball*-*comb*). Prime and target-distractor pairs were semantically unrelated. Interestingly, infants showed different target looking preference patterns after phonologically related prime pictures, compared to after phonologically unrelated primes. This suggests that infants implicitly named prime pictures---despite such pictures having been presented in silence---and that the phonology of the resulting labels interacted with subsequent target.

@chow2017spoken provided further evidence of semantic *and* phonological links in the early lexicon. The authors adapted the Visual World Paradigm from @huettig2007tug to explore 24- to 30-month-old toddlers' visual fixation patterns during a word recognition when presented with phonological and semantic distractors. In each trial, the authors presented participants with four semantically and phonologically pictures. Four seconds after pictures onset, a word-form was auditorily presented. The word did not refer to any of the pictures displayed on the screen, but was phonologically related to one of them (both labels shared phonological onset), and semantically related to another one of the pictures (both referents belonged to the same taxonomic category. For instance, participants might be presented with the pictures of a sandwich, a bus, a cat, and a dress. Then they would hear the carrier phrase "Look at the *bee*!". The authors registered participants fixations to the phonological and semantic distractors, and found evidence of a preference for the phonological distractor at earlier stages of the post-naming phase, and a preference for the semantic distractor at later stages of the trial. These results support a cascaded activation account of lexical access during the first stages of lexical development, paralleling previous findings in adults [@allopenna1998tracking].

One of the most interesting pieces of evidence in support of a cascaded account of lexical processing is provided by bilinguals. A critical property the bilingual lexicon is that its representations are accessed in a language-non selective way. Even during monolingual situations, bilinguals access their lexical representations in both of their languages [e.g., @thierry2007brain; @dunabeitia2009masked; @hoshino2008cognate; @strijkers2010tracking; @duyck2007visual]. This is reflected in bilinguals' performance in word recognition and production tasks. For instance, @marian1999activation presented Russian-English bilinguals with a series of trials that started with an auditory instruction like "Poloji marku nije krestika" ["Put the stamp below the cross"]. The instruction would be present in Russian (dominant language for most participants). The target object (*marku* [*stamp*], in this case) would be present in the screen, together with an object whose English label shared phonological onset with the target label (e.g., *mark*), and two unrelated distractors whose labels in Russian and English did not shared phonological relationship with the target label. In line with @allopenna1998tracking's results in monolingual adults, bilinguals fixated the cross-language phonological distractor significantly more than the unrelated distractors after hearing the target label. These findings suggest that bilinguals activated phonologically related word-forms in both Russian *and* English, which ultimately affected their overt visual exploration patterns during the task.

Language non-selectivity in the bilingual lexicon has also been shown to affect top-down lexical access during word production. @costa2000cognate presented Catalan-Spanish bilinguals with a series of pictures of familiar objects. For each object, participants were asked to name it in Catalan (their dominant language). Unbeknownst to participants, the authors manipulated the phonological overlap between the pictures' labels in Catalan and their translations in Spanish. In half of the trials, the labels associated to the pictures were cognates, that is, phonologically similar translation equivalents (e.g., *gat*-*gat* [*cat*]) in Catalan and Spanish, whereas in the other half of the trials the labels were non-cognates (e.g., *taula*-*mesa* [*table*]). Participants named pictures faster in cognate trials than in non-cognate trials. Critically, Spanish monolinguals who completed the same task, naming the pictures in Spanish, showed equivalent naming times in both conditions. These results reveal that bilinguals activated their Spanish phonology, despite performing the naming task exclusively in Catalan: the visual recognition of the presented pictures led to the activation of its associated phonological forms in both languages in parallel, which influenced the subsequent the dynamics of word production. This differential processing of cognates and non-cognates has generated a large body of literature providing support for the language-non selective account of lexical access [e.g., @degrootandnas1992; @schwartzkroll2007; @dijkstragrainger1999; @dufour1995matching; @schwartz2007reading; @dijkstra2010cross].

Further evidence in favour of the parallel retrieval of phonological forms during bilingual word production is given by implicit naming paradigms. Using an adaptation of @mani2010infant's task, @von2014bilinguals found that bilingual adults generate implicit labels in both labels for visually fixated pictures. The authors presented German-English bilinguals with 120 prime-target pairs. Primes were presented as familiar pictures in silence. After prime picture offset, target words were presented auditorily. Participants' N400 ERP components were recorded from target word onset. The authors manipulated the phonological relationship between the prime and target word-forms within participants' L1 and across L1 and L2. When prime and target were phonologically identical (*Affe*-*Affe*, German for *monkey*), or similar (*Fahne*-*Sahne*, German for *flag* and *ice-cream*, respectively) within L1, participants' showed a reduced N400 amplitude, compared to when prime and target were phonologically unrelated (*Messer*-*Seil*, German for *knife* and *rope*). Critically, a similar effect was found when prime and target were phonologically related through translation (*Rustsche*-*Kleid*, German for *slide* and *dress*). This suggests that participants activated prime labels in both languages in parallel, and that both labels impacted the dynamics of target words recognition.

Phonological priming paradigms have also revealed language non-selectivity in the initial bilingual lexicon [@poarch2012cross; @bosma2020cognate; @singh2014one; @jardak2019labels; @von2019impact]. For instance, @von2012language found evidence of cross-language phonological priming in 21- to 42-months-old children learning German and English. At the beginning of each trial, the authors auditorily presented an English prime word embedded in a carrier phrase. Then, the target label was also auditorily presented in German. Finally, the target and distractor pictures were presented side-by-side. In some trials, the auditorily presented prime and target labels were phonologically related (e.g., *slide*-*Klide* [*dress*]). In some other trials, both labels were phonologically related through translation: the auditorily presented prime label did *not* overlap with the target label (e.g., *leg*-*Stein* [*stone*]), but its translation in German did [*Bein*]. In the rest of the trials prime and target labels were phonologically unrelated in both languages. The authors found a facilitation effect of cross-language priming when both prime and target auditory labels overlapped phonologically, as revealed by a stronger target picture looking preference, as compared to that of unrelated trials. Interestingly, participants showed a weaker target preference in priming through translation trials.

Using a similar task, @floccia2020translation provided conflicting results to those of @von2012language. The authors tested 27-month-old simultaneous bilinguals in a cross-language priming paradigm. The two languages participants were learning were English, and an Additional Language (either Cantonese, Dutch, French, German, Greek, Italian, Mandarin, Polish, Portuguese, or Spanish). A prime word was embedded at the end of a carrier sentence that participants listened to. Then, participants were presented with the auditory label of the target word, and two pictures were shown side-by-side, the target picture, and a distractor picture. In Study 1, the target word was the translation equivalent of the prime word (*cheese*-*fromage* for an English-French bilingual). In other trials, prime and target words were unrelated (e.g., *sock*-*fromage*). There was no phonological overlap at onset between prime and target or distractor labels in either of the two conditions. Participants were presented with two blocks of related and unrelated trials. In one of the blocks, participants were presented with English prime words and Home Language target words (*sock*-*fromage*), and in the other block they were presented with Additional Language prime words and English target words (*fromage*-*sock*). The authors found a cross-linguistic priming effect, in which participants showed a stronger preference for the target picture in the post-naming phase in related trials, as compared to unrelated trials. This effect was found regardless of the language of the prime (English or Home Language), or the dominance of the language of the prime (i.e., whether the prime word belonged to the dominant language). Target preference exceeded chance level in both conditions, pointing to successful word recognition regardless of the experimental manipulation.

In a second study, @floccia2020translation, the authors tested an equivalent group of bilinguals in a cross-linguistic semantic priming task, in which participants were primed with words in English, and tested with semantically related words in the Additional Language (or vice versa). This time, prime and target words were not translation equivalents, but rather words whose referents shared semantic features. Results paralleled those from their first study. The authors found a strong semantic priming effect across languages, regardless of the language of the prime or the dominance of the language of the prime. Overall, these results suggest that bilinguals recognised target words in both conditions, and benefited from a cross-linguistic facilitation effect from translation equivalents. The symmetry of such effect across both languages suggests that participants lexical access occurred in parallel for the two members of translation equivalents. As the authors noted, the stimuli set of their second study contained multiple cognates due to design constraints, especially in languages with a larger degree of phonological overlap across translation pairs (e.g., English and Dutch). The authors run *post hoc* analyses exploring the effect of cognateness on the cross-language priming effects found, which showed no evidence of a cognateness effect on target preference. However, the cross-language effects reported in this second study might still reflect the confounding influence of cognateness, as the resulting effect might have been generated by an interaction by the phonological and semantic overlap between cognate primes and targets.

Previous studies had reported similar cross-language priming effects. The conclusion that such effects occurred at the lexical level, as opposed to being the result of mere acoustic similarity, was nonetheless unwarranted. In studies where the auditorily presented prime label was phonologically overlapped with the target label [e.g., @marian1999activation], priming effects might be explained by the activation of the phonological segments of the prime label, which facilitated the recognition of the target label, which shared such phonological segments. The priming-through-translation results by @von2012language and @floccia2020translation demonstrate that cross-language activation in the developing lexicon occurs at the lexical level. Since participants' target recognition was modulated by the non-presented prime translation, such an inhibitory effect must stem at the lexical level. The fact that phonological priming facilitated the recognition of the target words (both labels belong to different languages), and that phonological priming through translation interfered target recognition (translation and target belong to the same language) points to cross-language connections playing an excitatory role, and to within-language connections playing an inhibitory role.

The priming-through-translation paradigm provides a convenient framework to investigate parallel activation in the bilingual lexicon, but its design prevents strong interpretations about the extent to which parallel activation is the result of the experimental manipulations of interest. The presentation of primes and targets in both languages in the same task---or even within the same trial---presents participants with a bilingual context that might lead to a higher baseline activation of lexical representations of both languages, contrary to a context in which parallel activation results from genuine cross-language activation [@grosjean1997bilingual]. Another limitation of priming-through-translation studies is that, when prime words are auditorily presented, the task constrains the naming context to one of the languages. The consequence is a potential asymmetry between the activation dynamics of the lexical representation of both labels: while the participant has heard one label, the activation of its translation must spread across semantic links. This might lead to a lower strength, and higher latency of the activation of the translation.

In the present study, we put the hypothesis that bilingual infants activate both languages in parallel to a more severe test. We used an extension of the implicit naming paradigm by @mani2010infant, in which participants were primed with a picture presented in silence. This provides participants with an exclusively monolingual context in which they perform the task in only one of their languages. If differences in task performance are found associated with covert manipulation of cross-linguistic properties of the stimuli, such differences may be more conclusively linked to the parallel activation of lexical representations in participants' lexicons. By using an implicit naming paradigm, the activation of the lexical representation of the prime may occur in both languages as the result of lexical retrieval after visual recognition of the prime picture, since participants are not presented with any auditory prime label. This contrasts with priming through-translation tasks, in which the prime label in one of the languages is directly activated by auditory word recognition, while the label of its translation is expected to be activated as the result of cascaded activation spreading through their semantic association. This may lead to an asymmetry in the time course of the activation of both labels, which might in turn affect the dynamics of phonological priming in subsequent word recognition.

In summary, the present study was aimed at investigating the developmental trajectories of word recognition in bilingualism, and how (if at all) language non-selectivity shapes them. Using an adaptation of [@mani2010infant], we presented 20- to 32-months-old toddlers with a cross-linguistic phonological priming paradigm. We manipulated both the phonological overlap between the prime word and target word, and the cognate status of the prime word, to test the hypothesis that bilingual toddlers implicitly name familiar pictures in both languages. If so, phonological priming effect should be larger when the prime label in both languages shares phonological overlap with the target words (*Related/Cognate* condition), compared to when only the label of the prime in one language overlaps phonologically with the target (*Related/Non-cognate*). We also also included a control *Unrelated* condition in which regardless of the cognate status of the prime word, neither of the prime labels in any language overlapped phonologically with the target word.

To test cross-language priming effects, we registered participants' time course of target picture looking in the three conditions. We predicted within-language priming effects, revealed as differences in target looking between *Unrelated* and *Related/Non-cognate* trials, as previously reported [@von2012language, @mani2010infant, and @mani2011phonological]. We further predicted cross-language priming effects, revealed as differences in target looking between *Related/Non-cognate* and *Related/Cognate* trials. 
This effect should be stronger in bilinguals, if not completely absent in monolinguals, who should be indifferent to the phonological properties of the stimuli in the other language.

Following previous studies associating the emergence of phonological links between lexical items to the size of children's vocabulary [@mani2011phonological], we also compared the explanatory power of two indices of receptive vocabulary size (dominant-language vocabulary size and total vocabulary size) against participants' age. We used parental reports of vocabulary checklist to estimate participants' vocabulary size. We computed vocabulary size in the dominant language (L1 vocabulary) as the proportion of words in the dominant language reported by their caregivers to be understood by the child in a vocabulary checklist, and the total vocabulary size as the proportion of words reported as understood by the child in *both* languages. If within-language or cross-language phonological associations in the lexicon emerge as a function of vocabulary growth, the size of the differences between conditions should larger in participants with larger vocabulary sizes. A model including participants' L1 or total vocabulary size as a predictor should then show better performance than a model that instead includes participants age. In the case of bilinguals, if the emergence of cross-language priming effects is associated with the growth of their vocabulary is *both* languages, a model including total vocabulary size as a predictor should perform better than a model that includes L1 vocabulary size. The opposite pattern would indicate that cross-language priming effects emerge as a function of the growth of L1 vocabulary.

Testing bilingual populations at early ages presents the limitation that the vocabulary size of participants' non-dominant language is relatively small, which substantially constrains the number of items that can be used to as stimuli in the task, as well as the number of trials that can be constructed to generate stimuli lists. In order to circumvent the problem of limited vocabulary knowledge in the non-dominant language, tested participants only in their dominant language, therefore directing our research to the question of how possessing a second language affects processing of the dominant one in bilingual toddlers [@costa2014does provide a fuller description of the advantages of such approach].

# Study 1

## Methods

All materials, data, and reproducible code can be found at the OSF ([https://osf.io/hy984/](https://osf.io/ckydb/)) and GitHub ([https://github.com/gongcastro/cognate-priming](https://github.com/gongcastro/cognate-priming)) repositories. This study was conducted according to guidelines laid down in the Declaration of Helsinki, and was approved by the Drug Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar, reference 2020/9080/I. Before every testing session, caregivers were asked to read and sign an informed consent form, and were given a token of appreciation at the end of it.

### Participants

```{r}
#| label: participants-numbers
participants_bcn <- filter(participants, location=="Barcelona")

n_participants_total <- nrow(distinct(participants_bcn, child_id))
n_sessions_total <- count(participants_bcn)

n_participants_lp <- participants_bcn |> 
	distinct(lp, child_id) |> 
	count(lp)

n_participants_lp_sex <- participants_bcn |> 
	distinct(lp, sex, child_id) |> 
	count(lp, sex)

n_participants_sessions <- participants_bcn |> 
	count(child_id, name = "n_sessions") |> 
	count(n_sessions) |> 
	group_split(n_sessions) |> 
	set_names(paste0("session_", 1:3)) |> 
	map("n")

n_sessions_lp <- participants_bcn |> 
	summarise(across(age, lst(mean, sd, min, max)), n = n(), .by = c(lp)) |> 
	mutate(across(age_mean:age_max, \(x) round(x, 2))) |> 
	group_split(lp) |> 
	set_names(janitor::make_clean_names(unique(participants_bcn$lp)))

n_sessions_dominance_lp <- count(participants_bcn, lp, test_language) |> 
	group_split(test_language) |> 
	set_names(c("catalan", "spanish")) |> 
	map(group_split, lp)

n_sessions_lp <- participants_bcn |> 
	count(lp) |> 
	group_split(lp) |> 
	set_names(c("monolingual", "bilingual")) |> 
	map("n")
```

We collected data from `r n_participants_total` children living in the Metropolitan Area of Barcelona (Spain), tested at the Laboratori de Recerca en Infància at the Universitat Pompeu Fabra. Families were recruited from maternity rooms in private hospitals and social media, and contacted via phone when the child's age spanned between 20 and 32 months. From the `r n_participants_total` children that participated, `r n_participants_sessions$session_1` participated once, `r n_participants_sessions$session_2` participated twice, and `r n_participants_sessions$session_3` participated three times. Recurrent participants were tested with at least XXX months of difference. We gathered a total of `r n_sessions_total$n` testing sessions. Participants were divided into monolinguals and bilinguals based on their relative degree of exposure to Catalan and Spanish, estimated using the Language Exposure Questionnaire [LEQ, @bosch2001evidence]. We categorised participants as monolingual if exposed to more than 80% or more of the time to their dominant language, and as bilingual otherwise. `r english::Words(n_participants_lp$n[1])` of the participants were categorised as monolinguals (`r n_participants_lp_sex$n[1]` female, `r n_participants_lp_sex$n[2]` male) and `r n_participants_lp$n[2]` as Catalan/Spanish bilinguals (`r n_participants_lp_sex$n[3]` female, `r n_participants_lp_sex$n[4]` male) (see @tbl-participants for a detailed summary of participants' age and language profile). Participants' vision was normal, none used glasses or any other type of vision corrector.

```{r tbl-participants}
#| label: tbl-participants
#| tbl-cap: "Demographic and linguistic profile of testing sessions."
participants_bcn |> 
	unnest_wider(doe) |> 
	select(-matches("others")) |> 
	arrange(lp, test_language) |> 
	summarise(n = n(),
			  n_participants = n_distinct(child_id),
			  across(age, lst(mean, sd, min, max)),
			  across(matches("doe_"), lst(mean, sd)),
			  # age_vctr = list(age),
			  .by = c(lp, test_language)) |> 
	gt(rowname_col = "test_language",
	   groupname_col = "lp") |> 
	# gt_plt_dist(age_vctr, type = "histogram") |>
	fmt_number(is.numeric, decimals = 1) |> 
	fmt_integer(c(n, n_participants)) |> 
	fmt_number(matches("doe"), decimals = 1, scale_by = 100) |> 
	cols_merge(c(age_min, age_max), 
			   pattern = "{1} ({2})") |> 
	cols_merge(c(age_mean, age_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(doe_catalan_mean, doe_catalan_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(doe_english_mean, doe_english_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(doe_spanish_mean, doe_spanish_sd), 
			   pattern = "{1} ({2})") |> 
	tab_spanner("Age (months)", matches("age")) |> 
	tab_spanner("Catalan", matches("catalan")) |> 
	tab_spanner("English", matches("english")) |> 
	tab_spanner("Spanish", matches("spanish")) |> 
	tab_spanner("Degree of Exposure (%)", matches("doe")) |> 
	cols_label(n = "N sessions",
			   n_participants = "N participants",
			   age_mean = "M (SD)",
			   age_min = "Range",
			   # age_vctr = "Distribution",
			   doe_catalan_mean = "M (SD)",
			   doe_english_mean = "M (SD)",
			   doe_spanish_mean = "M (SD)") |> 
	tab_style(cell_text(style = "italic"),
			  cells_column_labels()) |> 
	tab_style(cell_text(style = "italic"),
			  cells_column_spanners())

# col_names <- c("Group",
# 			   "Language",
# 			   "Sessions", 
# 			   "Participants",
# 			   "M (SD)",
# 			   "Range",
# 			   "M (SD)",
# 			   "M (SD)",
# 			   "M (SD)")
# participants_bcn |> 
# 	unnest_wider(doe) |> 
# 	select(-matches("others")) |> 
# 	arrange(lp, test_language) |> 
# 	summarise(n = n(),
# 			  n_participants = n_distinct(child_id),
# 			  across(age, lst(mean, sd, min, max)),
# 			  across(matches("doe_"), lst(mean, sd)),
# 			  # age_vctr = list(age),
# 			  .by = c(lp, test_language)) |> 
# 	mutate(age_mean = sprintf("%.1f (%.1f)", age_mean, age_sd),
# 		   age_range = sprintf("%.1f-%.1f", age_min, age_max),
# 		   doe_catalan = sprintf("%.1f (%.1f)", doe_catalan_mean*100, doe_catalan_sd*100),
# 		   doe_spanish = sprintf("%.1f (%.1f)", doe_spanish_mean*100, doe_spanish_sd*100),
# 		   doe_english = sprintf("%.1f (%.1f)", doe_english_mean*100, doe_english_sd*100)) |> 
# 	select(lp:age_mean, age_range, doe_catalan, doe_spanish, doe_english) |> 
# 	kable(digits = 1, 
# 		  format = "latex",
# 		  booktabs = TRUE,
# 		  col.names = col_names,
# 		  align = "llrrrrrrrrrrrr"
# 	) |> 
# add_header_above(c(
# 	" " = 4,
# 	"Age (months)" = 2,
# 	"Degree of exposure (%)" = 8
# )) |> 
# 	row_spec(nrow(tbl_data) - 1,
# 			 hline_after = TRUE
# 	) 
# 	add_footnote("For participants exposed to Catalan and Spanish exclusively, the proportion of Catalan exposure shown in the table is complementary to the degree of exposue to Spanish. For participants exposed to a third languages this proportion is not complementary unless one adds the degree of exposed to the third language, which never exceeded 10\\%. Participants with 100\\%, 0\\%, and 50\\% Catalan exposure would be traditionaly classified as Catalan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively.", notation = "none", escape = FALSE)
```

### Vocabulary size

```{r vocab-values}
vocabulary_bcn <- vocabulary |> 
	inner_join(select(participants_bcn, session_id, lp, age_group),
			   by = join_by(session_id))

vocab_means <- vocabulary_bcn |> 
	summarise(across(c(total_prop, l1_prop), lst(mean, sd)),
			  n = n(),
			  .by = c(age_group, lp)) %>%
	mutate(across(matches("prop"), scales::percent)) %>%
	split(paste0("age_", gsub(" months", "", .$age_group))) |> 
	map(\(x) split(x, tolower(x$lp)))

n_imputed <- table(vocabulary_bcn$is_imputed)[2]
prop_imputed <- scales::percent(n_imputed/nrow(vocabulary_bcn))
n_pool <- nrow(bvq_data$vocabulary)
```

We collected vocabulary data using parental responses to the Barcelona Vocabulary Inventory [BVQ, @garcia-castro2023bvq], an online vocabulary checklist inspired in several adaptations of the the Communicative Developmental Inventory [CDI, @fenson1994variability] developed to assess the vocabulary size of Catalan-Spanish bilingual toddlers. Families received a link to the BVQ immediately after each experimental session, and were given two weeks to fill it. We calculated two measures of receptive vocabulary size for each participant: *L1 vocabulary size* (proportion of words  reported as acquired in the checklist of the dominant language), and *total vocabulary size* (proportion of the words in both checklists reported as acquired). `r n_imputed` (`r prop_imputed`) Families failed to provide a complete response to the BVQ within the two-week time limit, or did not provide a successful response to the questionnaire. For missing questionnaire responses, we imputed the vocabulary size of the participant using single imputation, using the vocabulary size scores of a pool of `r n_pool` additional participants for which a successful response for the questionnaire had been gathered. We used participants age in months and their language profile (monolingual/bilingual) as predictors. We used the `mice` R package [@van2011mice] to perform imputation using the Bayesian linear regression method.

Overall, there was a substantial increase in both total and L1 vocabulary sizes associated to age (see @fig-vocabulary). Monolinguals showed overall higher L1 vocabulary sizes than monolinguals: they were reported to understand an average of `r vocab_means$age_21$monolingual$l1_prop_mean` (*SD* = `r vocab_means$age_21$monolingual$l1_prop_sd`) of the words at around 21 months, `r vocab_means$age_25$monolingual$l1_prop_mean` (*SD* = `r vocab_means$age_25$monolingual$l1_prop_sd`) at 25 months, and `r vocab_means$age_30$monolingual$l1_prop_mean` (*SD* = `r vocab_means$age_30$monolingual$l1_prop_sd`) at 30 months, while bilinguals  were reported to understand an average of `r vocab_means$age_21$bilingual$l1_prop_mean` (*SD* = `r vocab_means$age_21$bilingual$l1_prop_sd`) of the words at around 21 months, `r vocab_means$age_25$bilingual$l1_prop_mean` (*SD* = `r vocab_means$age_25$bilingual$l1_prop_sd`) at 25 months, and `r vocab_means$age_30$bilingual$l1_prop_mean` (*SD* = `r vocab_means$age_30$bilingual$l1_prop_sd`) at 30 months. Total vocabulary sizes showed the opposite pattern, with monolinguals infants being reported to understand `r vocab_means$age_21$monolingual$total_prop_mean` (*SD* = `r vocab_means$age_21$monolingual$total_prop_sd`) of the words at around 21 months, `r vocab_means$age_25$monolingual$total_prop_mean` (*SD* = `r vocab_means$age_25$monolingual$total_prop_sd`) at 25 months, and `r vocab_means$age_30$monolingual$total_prop_mean` (*SD* = `r vocab_means$age_30$monolingual$total_prop_sd`) at 30 months, and bilinguals being reported to understand `r vocab_means$age_21$bilingual$total_prop_mean` (*SD* = `r vocab_means$age_21$bilingual$total_prop_sd`) of the words at around 21 months, `r vocab_means$age_25$bilingual$total_prop_mean` (*SD* = `r vocab_means$age_25$bilingual$total_prop_sd`) at 25 months, and `r vocab_means$age_30$bilingual$total_prop_mean` (*SD* = `r vocab_means$age_30$bilingual$total_prop_sd`) at 30 months.

>>>>

---
---

```{r stimuli-values}
n_words <- stimuli |>
	filter(test_language=="English") |> 
	select(prime, target, distractor) |> 
	unlist() |> 
	n_distinct()
```

We used `r n_words` distinct words to create the stimuli lists. We created four stimuli lists of 32 trials, each trial involving a prime-target-distractor triplet. Words played a role as primes *or* targets/distractors across the three lists in their corresponding language. For instance, the word *boot* appeared as *prime* in the three lists, but never as *target* or *distractor*; the word *bird* appeared as *target* and *distractor* across the three lists, but never as a prime. Target-distractor pairings were held constant across the three lists in each language. For instance, in all Catalan lists the word *bird* was paired with the word *milk*. Target-distractor pairings were also yoked, so that each member of the same target-distractor pair appeared once as target and once as distractor in each list. For instance, the *bird*-*milk* pair appeared twice in each of the four lists: once with *bird* as target and *milk* as distractor, and once with *milk* as target and *bird* as distractor. This counterbalancing avoided participants encountering looking at the target word guided solely by that word having being named in a previous trial. Finally, prime words appeared only once in each list: each target-distractor pair was associated with a different prime word in both appearances. Across lists, the same prime word was presented alongside a different target-distractor pair. For instance, the prime word *boot* was presented with the *bird*-*milk* target-distractor pair in the first list, with the *bucket*-*mouse* pair the second list, with *balloon*-*snail* in the third list, and with *bread*-*fish* in the fourth list. To avoid competition between target and distractor pictures, semantically related target-distractor pairs were perceptually distinct [@floccia2020translation; @arias2009lexical]. The order of the trials was randomised across experimental sessions, so that each time a participant was tested, the order in which the prime-target-distractor was presented was random. Each participant was randomly assigned to one of the three lists in the corresponding (dominant) language, and always the same list across their experimental sessions in the case of a recurrent participant. The position of the target and distractor pictures (right or left) for each target-distractor pair was alternated, so that in one list the target would appear on the left, in another list it would appear on the right, and so on.

The speaker was presented with a list of words in Catalan. The order of the words was pseudo-randomised, and each word was produced three times in a row before moving to the next word in the list. After going through all the words in the list, the speaker went through the word list again generating three tokens for each word, now in an inverse order (from bottom of the list to the top). We then repeated the same procedure for the list of Spanish words. The resulting audios were manually chunked into individual word-forms. For each of the six tokens produced for each word, the most adequate was selected for further processing. The audios were then transformed to stereo by duplicating them into two channels, denoised, and finally normalised. The mean duration of the final audios was `r round(mean(stimuli$duration[stimuli$test_language=="Catalan"]), 2)` (*SD* = `r round(sd(stimuli$duration[stimuli$test_language=="Catalan"]), 2)`) and `r round(mean(stimuli$duration[stimuli$test_language=="Spanish"]), 2)` (*SD* = `r round(sd(stimuli$duration[stimuli$test_language=="Spanish"]), 2)`) seconds for the Catalan and Spanish lists.
